---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# iSEEde

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/iSEE/iSEEde)](https://github.com/iSEE/iSEEde/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/iSEE/iSEEde)](https://github.com/iSEE/iSEEde/pulls)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check-bioc](https://github.com/iSEE/iSEEde/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/iSEE/iSEEde/actions)
[![Codecov test coverage](https://codecov.io/gh/iSEE/iSEEde/branch/main/graph/badge.svg)](https://app.codecov.io/gh/iSEE/iSEEde?branch=main)
<!-- badges: end -->

The goal of `iSEEde` is to provide panels to facilitate the interactive visualisation of differential expression results in `r BiocStyle::Biocpkg("iSEE")` applications.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `iSEEde` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("iSEEde")
```

And the development version from [GitHub](https://github.com/iSEE/iSEEde) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("iSEE/iSEEde")
```

## Example

This is a basic example which shows you how to load the package:

```{r example, message=FALSE, warning=FALSE, eval = requireNamespace('iSEEde')}
library("iSEEde")
library("airway")
library("DESeq2")
library("iSEE")

# Example data ----

data("airway")
airway$dex <- relevel(airway$dex, "untrt")

dds <- DESeqDataSet(airway, ~ 0 + dex + cell)

dds <- DESeq(dds)
res_deseq2 <- results(dds, contrast = list("dextrt", "dexuntrt"))

# iSEE ---

airway <- embedContrastResults(res_deseq2, airway, name = "dex: trt vs untrt")

app <- iSEE(airway, initial = list(
  DETable(ContrastName="dex: trt vs untrt", HiddenColumns = c("baseMean", 
    "lfcSE", "stat"), PanelWidth = 4L),
  VolcanoPlot(ContrastName="dex: trt vs untrt", PanelWidth = 4L),
  MAPlot(ContrastName="dex: trt vs untrt", PanelWidth = 4L)
))

if (interactive()) {
    shiny::runApp(app)
}
```

## Citation

Below is the citation output from using `citation('iSEEde')` in R. Please
run this yourself to check for any updates on how to cite __iSEEde__.

```{r 'citation', eval = requireNamespace('iSEEde')}
print(citation("iSEEde"), bibtex = TRUE)
```

Please note that the `iSEEde` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `iSEEde` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://iSEE.github.io/iSEEde) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

## Code of Conduct

Please note that the iSEEde project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.
