---
title: "Supported differential expression methods"
author: 
  - name: Kevin Rue-Albrecht
    affiliation:
    - University of Oxford
    email: kevin.rue-albrecht@imm.ox.ac.uk
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('iSEEde')`"
vignette: >
  %\VignetteIndexEntry{Supported differential expression methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r, eval=!exists("SCREENSHOT"), include=FALSE}
SCREENSHOT <- function(x, ...) knitr::include_graphics(x)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    iSEEde = citation("iSEEde")[1]
)
```

# Differential expression results

## Formats

Differential expression results are generally reported as tables of statistics, including (log) fold-change, p-value, average expression, etc.

Those statistics being reported for individual features (e.g., genes), the `rowData()` component of `SummarizedExperiment()` objects provides a natural home for that information.
Specifically, `r BiocStyle::Biocpkg("iSEEde")` searches for differential expression results in `rowData(se)[["iSEEde"]]`.

The first challenge arises when differential expression statistics are computed only for a subset of features.
In that case, `r BiocStyle::Biocpkg("iSEEde")` populates missing information with `NA`.

The second challenge arises from the different names of columns used by individual differential expression methods to store differential expression common statistics.
To address this, `r BiocStyle::Biocpkg("iSEEde")` provides S4 classes creating a common interface to supported differential expression methods.

# Example data

We use the `?airway` data set.

We briefly adjust the reference level of the treatment factor to the untreated condition.

```{r, message=FALSE, warning=FALSE}
library("airway")
data("airway")
airway$dex <- relevel(airway$dex, "untrt")
```

# Supported methods

## Limma

```{r, message=FALSE, warning=FALSE}
library("edgeR")

counts <- assay(airway, "counts")
design <- model.matrix(~ 0 + dex + cell, data = colData(airway))

keep <- filterByExpr(counts, design)
v <- voom(counts[keep,], design, plot=FALSE)
fit <- lmFit(v, design)
contr <- makeContrasts("dextrt - dexuntrt", levels = colnames(coef(fit)))
tmp <- contrasts.fit(fit, contr)
tmp <- eBayes(tmp)
res_limma <- topTable(tmp, sort.by = "P", n = Inf)
head(res_limma)
```

To store this first set of differential expression results, we initialise a `DataFrame()` nested in the `"iSEEde"` column of `rowData(airway)`.
The constructor `iSEELimmaResults()` is given the rownames of the full `airway` object, to match the dimensions of the dataset and fill missing information with `NA`.
The resulting `DataFrame`-like object is then stored as a nested column, using a name that identifies the contrast.

We embed this set of differential expression results in the `airway` object using the `embedResults()` method.
Specifically, the results are nested as a `DataFrame`-like object in `rowData(airway)[["iSEEde"]]`.

```{r}
library(iSEEde)
airway <- embedResults(
  iSEELimmaResults(res_limma, row.names = rownames(airway)),
  airway,
  name = "Limma-Voom")
rowData(airway)
```

## DESeq2

```{r, message=FALSE, warning=FALSE}
library("DESeq2")

dds <- DESeqDataSet(airway, ~ 0 + dex + cell)

dds <- DESeq(dds)
res_deseq2 <- results(dds, contrast = list("dextrt", "dexuntrt"))
head(res_deseq2)
```

We add this set of differential expression results as a `DataFrame`-like object nested in a column of `rowData(airway)[["iSEEde"]]`.

```{r}
airway <- embedResults(
  iSEEDESeq2Results(res_deseq2, row.names = rownames(airway)),
  airway,
  name = "DESeq2")
rowData(airway)
```

## edgeR

```{r, message=FALSE, warning=FALSE}
library("edgeR")

counts <- assay(airway, "counts")
design <- model.matrix(~ 0 + dex + cell, data = colData(airway))

fit <- glmFit(counts, design, dispersion=0.1)
lrt <- glmLRT(fit, contrast = c(-1, 1, 0, 0, 0))
res_edger <- topTags(lrt, n = Inf)
head(res_edger)
```

We add this set of differential expression results as a `DataFrame`-like object nested in a column of `rowData(airway)[["iSEEde"]]`.

```{r}
airway <- embedResults(
  iSEEedgeRResults(res_deseq2, row.names = rownames(airway)),
  airway,
  name = "edgeR")
rowData(airway)
```

# Live app

```{r}
library(iSEE)
app <- iSEE(airway, initial = list(
  VolcanoPlot(ContrastName="Limma-Voom", PanelWidth = 6L),
  MAPlot(ContrastName="Limma-Voom", PanelWidth = 6L),
  VolcanoPlot(ContrastName="DESeq2", PanelWidth = 6L),
  MAPlot(ContrastName="DESeq2", PanelWidth = 6L),
  VolcanoPlot(ContrastName="edgeR", PanelWidth = 6L),
  MAPlot(ContrastName="edgeR", PanelWidth = 6L)
))

if (interactive()) {
  shiny::runApp(app)
}
```

```{r, echo=FALSE, out.width="100%"}
SCREENSHOT("screenshots/volcano_plots.png", delay=20)
```

# Reproducibility

The `r Biocpkg("iSEEde")` package `r Citep(bib[["iSEEde"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("methods.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("methods.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```

<!-- Links -->

[scheme-wikipedia]: https://en.wikipedia.org/wiki/Uniform_Resource_Identifier#Syntax
[iana-uri]: https://www.iana.org/assignments/uri-schemes/uri-schemes.xhtml